Belowground sink limitation alters growth and carbon balance of *Eucalyptus* seedlings

### 2016-01-11 manuscript/manuscript_pve.Rmd
final thesis edits from MT

Interpreting limitations to plant growth requires understanding the balance between carbon (*C*) source and sink activity. This study used manipulations of soil volume to test how growth is coupled to physiology, *C* allocation, and sink activity in *Eucalyptus tereticornis* seedlings. We grew individual seedlings in a large range of container sizes and planted containers flush to the soil alongside naturally planted freely-rooted (free) seedlings. We developed a seedling growth model that utilized leaf photosynthesis rates (*A*~n~) to allocate daily *C* uptake towards mass growth of stems, leaves and roots. Reduced soil volume was expected to induce rapid negative effects on growth and physiology compared to free seedlings. It was hypothesized that the soil volume effect would be largest in the smallest containers, negatively impacting mass partitioning belowground. An accumulation of leaf non-structural carbohydrates, resulting from reduced belowground sink strength, was expected to correlate to reductions in photosynthetic capacity. We observed a negative effect of container volume on aboveground growth soon after the experiment started. Although growth was consistently different across soil volumes, dry mass partitioning to leaves, stems and roots was unchanged after 120 days. Photosynthetic capacity was significantly reduced in containers compared to free seedlings, and was related to both reduced leaf nitrogen content and starch accumulation. We then asked whether the observed reductions in *A*~n~ explained the observed differences in seedling biomass. We found that although belowground sink limitation resulted in down regulation of *A*~n~, these reductions were not large enough to explain observed growth responses. Thus, as *A*~n~ and growth were not tightly coordinated, excess photosynthetic *C* not attributed to biomass resulted in seedlings with soil volume restriction. These results highlight the need to further utilize mass balance approaches when evaluating plant C allocation and confirm that important feedbacks exist between belowground sink strength and leaf *C* uptake.

@@

\
(1)
$$P_i = L\left(\frac{C_\mathrm{day'i}~\sigma_\mathrm{s}}{\epsilon_\mathrm{c}}\right) - R$$
where *L* is total plant leaf area (m^2^), *C*~day,i~ is the predicted daily carbon assimilation (g C g mass^-1^ d^-1^,, explained below), $\sigma$~s~ is a self shading parameter (explained below), $\epsilon$~c~ is a biomass conversion efficiency parameter (g C g mass^-1^) and *R* is the mass based total respiration of all tissue components (g C d^-1^). Total respiration was calculated as

>>>>

Interpreting limitations to plant growth requires understanding the balance between carbon (*C*) source and sink activity. This study used manipulations of soil volume to test how growth is coupled to physiology, *C* allocation, and sink activity in *Eucalyptus tereticornis* seedlings. We grew individual seedlings in a large range of container sizes and planted containers flush to the soil alongside naturally planted freely-rooted (free) seedlings. We developed a seedling growth model that utilized leaf photosynthesis rates (*A*~n~) to allocate daily *C* uptake towards mass growth of stems, leaves and roots. Reduced soil volume was expected to induce rapid negative effects on growth and physiology compared to free seedlings. It was hypothesized that the soil volume effect would be largest in the smallest containers, negatively impacting mass partitioning belowground. An accumulation of leaf non-structural carbohydrates, resulting from reduced belowground sink strength, was expected to correlate to reductions in photosynthetic capacity. We observed a negative effect of container volume on aboveground growth soon after the experiment started. Although growth was consistently different across soil volumes, dry mass partitioning to leaves, stems and roots was unchanged after 120 days. Photosynthetic capacity was significantly reduced in containers compared to free seedlings, and was related to both reduced leaf nitrogen content and starch accumulation. We then asked whether the observed reductions in *A*~n~ explained the observed differences in seedling biomass. We found that although belowground sink limitation resulted in down regulation of *A*~n~, these reductions were not large enough to explain observed growth responses. Thus, as *A*~n~ and growth were not tightly coordinated, an unaccounted for pool of photosynthetic *C* resulted in seedlings with soil volume restriction. These results highlight the need to further utilize mass balance approaches when evaluating plant C allocation and confirm that important feedbacks exist between belowground sink strength and leaf *C* uptake.

@@

\
(1)
$$P_i = L\left(\frac{C_\mathrm{day'i}~\sigma_\mathrm{s}}{\epsilon_\mathrm{c}}\right) - R$$
where *L* is total plant leaf area (m^2^), *C*~day,i~ is the predicted daily carbon assimilation (g C g mass^-1^ d^-1^,, explained below), $\sigma$~s~ is a self-shading parameter (explained below), $\epsilon$~c~ is a biomass conversion efficiency parameter (g C g mass^-1^) and *R* is the mass based total respiration of all tissue components (g C d^-1^). Total respiration was calculated as
